Exploratory data structure comparisons: Three new visual tools based on Principal Component Analysis

### 2018-05-17 article/Old versions/MPS/PCADSC_article.tex
updated version
updates as outlined in MPS_reviewercomments_response 17MAJ2018

\documentclass[a4paper,14pt]{article}
\usepackage[danish]{babel}
\usepackage{authblk}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathrsfs}
%\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
\usepackage{apalike}
\usepackage[utf8x]{inputenc}
%\usepackage{nameref,hyperref}
% line numbers
%\usepackage[right]{lineno}
% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }
% color can be used to apply background shading to table cells only
\usepackage[table]{xcolor}
% array package and thick rules for tables
\usepackage{array}
% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

@@

\def\volumeyear{2017}

@@

\author[1]{}
%\author[1]{Anne Helby Petersen}
%\author[1]{Bo Markussen}
%\author[3]{Karl Bang Christensen}
%\affil[1]{Department of Public Health, University of Copenhagen, Copenhagen, Denmark}
%\affil[2]{Department of Mathematical Sciences, University of Copenhagen, Copenhagen, Denmark}

\maketitle

@@

\newpage

@@

\begin{abstract}
Researchers often encounter data that is divided into distinct subsets. This might be due to multi-center sampling frames, or to variations in instruments, questionnaire item ordering or mode of administration. Often, one wishes to combine such data into one and conduct only a single analysis, but this is only meaningful if the subsets are similar in structure. Common procedures for assessing whether such a data collapsing is allowed are ad hoc and require making full model assumptions. We present visual diagnostic tools for investigating differences in structure for two datasets containing different observations of the same variables. This approach, referred to as \textit{Principal Component Analysis-based Data Structure Comparisons} (PCADSC), requires no directional nor hierarchical assumptions on the data and consists of three plots based on Principal Component Analysis (PCA), that address differences in the structures of the covariance matrices of the two datasets to yield an  understanding of where differences occur and why so. The functionality of the PCADSC tools is presented by use of data from the European Social Survey on psychological well-being in three European countries, namely Denmark, Sweden and Bulgaria. Here, we find that the concepts of psychological well-being are not the same when comparing Denmark and Bulgaria, and thus a comparison of e.g. mean scores in different well-being aspects is not meaningful. However, when comparing the two Scandinavian countries, Denmark and Sweden, we find very similar data structures and thus very similar concepts of well-being. Therefore, inter-country comparisons are warranted for these two countries.
\end{abstract}

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\date{}

@@

\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

@@

Data comparability is a classical topic in statistics and a recurring one in psychometrics. Often, psychometric data will be collected in such a way that it is essentially divided into several subsets whose comparability (and thus collapsibility) needs to be assessed empirically. This happens e.g. when data are collected across several centers (or countries) or when different versions of an instrument, the mode of administration or the order of survey questionnaire items are applied. All of these practices are essential parts of the psychometric methodology, as it allows us to simultaneously build on existing methods and answer new, empirical questions. However, if we do not address the question of collapsibility among the data subsets arriving from e.g. different countries, we risk conducting analyses whose most fundamental assumptions are not satisfied.

Sophisticated methods for addressing this question are available when we are willing to assume a statistical model. But this places the effort of assessing data comparability very late in the data analysis process and it makes the comparability assessment ad-hoc, as it essentially relies on the appropriateness of the modeling choices, which again depends on the structure of the data. The use of (parametric) models does thus not constitute a %valid
general data structure comparison method, but rather a fitted-model comparison method. It addresses the interplay between the model and the data, not the data alone. However, useful tools for initial, exploratory investigations of data comparability that help researchers identify potential problems or challenges early in the study design development are mostly absent. What is needed is a procedure that compares differences in overall data structures in two subsets of a dataset without assuming neither directional nor hierarchical relationships between the variables. Simple methods like variable-by-variable tests of distributional differences suffer from the drawback that they only address marginal differences and ignore the interplay between variables, but entry-by-entry comparisons of two empirical correlation matrices quickly become unmanageable as the number of variables increase.

@@

We propose three visual tools for this task, which we will refer to collectively as Principal Component Analysis-based Data Structure Comparisons (PCADSC). These methods use principal component decomposition of the empirical covariance matrix in the two sub sets to create intuitive visualizations of data structure differences. The proposed tools are implemented in the \texttt{R} package \texttt{PCADSC} \cite{PCADSC}.

@@

\subsection*{A brief introduction to principal component analysis}
\label{sec:introPCA}
Consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ of $d$ variables, let $\bar{x} = {(\bar{x}_1, ..., \bar{x}_d)}^\top = \frac{1}{n} \sum_{i=1}^n x_n$ denote their averages, and let $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) {(x_i-\bar{x})}^\top \in \RR^{d \times d}$ denote the empirical covariance matrix of the full data matrix, $X$. We assume in the following that all variables of $X$ have a numerical interpretation, e.g. by being continuous or ordinal categorical. For a given $q \leq d$, principal component analysis (PCA) is a tool for finding a new representation of the dataset of dimension $q$ such that the least possible amount of information is lost. More specifically, we wish to minimize the loss when looking at a projection of $X$ onto a $q$-dimensional space, rather than the original $d$-dimensional one. Formally, we define the \emph{rank-q-reconstruction error} as the minimal squared error that is achievable by linear subspaces $K_q \subset \RR^d$ of dimension $q < d$, that is

\begin{equation*}
\hat{K}_q = \text{span}\{\eta_1,\dotsc,\eta_q\},
\end{equation*}
where $\eta_1, ..., \eta_q$ are the first $q$ eigenvectors of the empirical covariance matrix, $S$, as ordered by the size of their associated eigenvalues, $\lambda_1, \dotsc, \lambda_d$. In the PCA framework, we refer to the eigenvectors as \textit{loadings}. The eigenvalues may be understood as \textit{variance components}, as the sum of the marginal empirical variances is preserved under eigenvalue decomposition, that is,
\begin{equation*}
\text{trace}(S) = \sum_{j=1}^d \hat{V}(x_j) = \sum_{j=1}^d \hat{V}(\eta_j^\top X^\top) = \sum_{j=1}^d \lambda_j,
\end{equation*}
where $x_j \in \RR^n$ denotes the $j$th variable of $X$ and $\hat{V}$ is the empirical variance function. This again emphasizes that we do not change the covariance structure of a dataset when performing PCA; we merely use linear algebra to make it easier to describe. And as eigenvalues are uniquely defined, and eigenvectors are uniquely defined up to a change of sign whenever the eigenvalues are %different,
distinct, the representation hereby obtained is a valid object of inference. If $n < d$ or if the covariance matrix does not have full rank, it is possible to obtain non-unique eigenvalues, e.g.\ $\lambda_i=\lambda_{i+1}=\dotsm=\lambda_j$, but even in this case, the associated eigenvectors $\eta_i,\eta_{i+1},\dotsc,\eta_j$ are uniquely defined up to a common rotation.

@@

All in all, PCA qualifies as an appealing first step in structural comparisons of two datasets containing the same variables, and especially the loadings and variance components are meaningful and interesting objects to compare across such different datasets. Usually, in PCA, main interest lies in the \textit{scores}, i.e. the projections $\eta_j^\top (x_i - \bar{x})$ of the observations onto the loadings, but the variation components and the accompanying loadings also describe aspects of the data. If two different datasets with the same variables, but different samples of observations, have similar loadings, then the variables can be argued to measure the same underlying quantities in both datasets. Similar loadings can occur in situations where the two sets of scores are different, e.g.\ if the two datasets come from two different populations of subjects. On the other hand, if the loading patterns are different in the two datasets, then this indicates that the variable interplay differs in the two data situations, and hence it would be criticizable to use these variables for comparisons across the two datasets.

The tools presented in this paper are all based on comparing the PCA results across two different datasets that contain the same variables. We denote these two datasets by $X_1$ and $X_2$, respectively, with $X_1$ consisting of $n_1$ observations of $d$ variables $x_{11},\dotsc,x_{1 n_1} \in \RR^d$, and similarly, $X_2$ consisting of $n_2$ observations of the apparently same $d$ variables, $x_{21},\dotsc,x_{2 n_2} \in \RR^d$. %
Furthermore, let $X$ be the combined dataset consisting of all $n=n_1+n_2$ observations. For these %two
three datasets, we complete the following steps:
\begin{enumerate}
\item Within each dataset, standardize each of the variables to have mean zero and unit standard deviation. Let $\tilde{x}_{ij} \in \RR^d$ for $i=1,2$ and $j=1,\dotsc,n_i$ and $\tilde{x}_j \in \RR^d$ for $j=1,\dotsc,n$ be the standardized datasets.
\item Form the principal component analyses
\begin{align*}
S_i &= \frac{1}{n_i-1} \sum_{j=1}^{n_i} \tilde{x}_{ij} \tilde{x}_{ij}^\top = \sum_{k=1}^d \lambda_{ik} \eta_{ik} \eta_{ik}^\top \quad \text{for $i=1,2$,} \\
S &= \frac{1}{n-1} \sum_{j=1}^n \tilde{x}_j \tilde{x}_j^\top = \sum_{k=1}^d \lambda_k \eta_k \eta_k^\top,
\end{align*}
thereby obtaining loadings $\eta_{1k}, \eta_{2k}, \eta_k$ and variance components $\lambda_{1k}, \lambda_{2k}, \lambda_k$ for $k=1,\dotsc,d$.
\end{enumerate}
The hereby obtained PCA decompositions of the correlation matrices can then be compared. We present three diagnostic plots that are designed to shine a light on different types and levels of data structure differences. These three plots are:
\begin{description}
\item[The CE plot:] The CE (cumulative eigenvalue) plot can be used to illustrate differences in variance components, that is, in the relative importance of the directions identified by the PCA. The CE plot is accompanied by two permutation-tests, testing the hypothesis of no difference in variance components.
\item[The angle plot:] The angle plot compares both loadings and variance components at once and it can be used to understand the information loss if the data structure of one dataset is superimposed on the other, thereby revealing which principal components (i.e. loading and variance component pairs) that are most similar and most different across the two datasets
\item[The chroma plot:] The chroma plot is primarily an illustration of the loading patterns and it targets the question of how the roles of the original variables are different between the two datasets, thus leading the data structure comparison question back to its original, empirical context.
\end{description}
For the deepest understanding of the data structure differences in two datasets, we suggest using all three steps in the above order.

Note that the standardization implies that the diagonal elements of $S$, $S_1$, and $S_2$ all equals 1, and thus also that
\begin{equation*}
\sum_{k=1}^d \lambda_k = \sum_{k=1}^d \lambda_{1k} = \sum_{k=1}^d \lambda_{2k} =  d.
\end{equation*}
This identity will simplify some expressions below.

@@

In \textit{Model 1} the observed variables $v_3$ and $v_4$ are caused by a common, unobserved variable $X_1$. Further details on the models and the simulated data are available in the supplementary material, including covariance matrices for the two models. Further details about data simulation is also available in the supplementary material. In dataset A, all observations stem from the same, underlying data structure and thus, the PCADSC plots should point towards no notable differences in data structures. In dataset B, on the other hand, there are two fundamentally different, underlying data structures and therefore, the PCADSC plots should illustrate this lack of homogeneity in data structures.

\begin{figure}[!h]
\caption{[FIGURE 1 HERE]}
% Graphical representations of models used for simulating example data. Square nodes represent observed variables, while rounded nodes represent latent variables. Arrows are used to illustrate causal links.
\label{fig:simGraphs}
\end{figure}

\subsubsection*{The cumulative eigenvalue plot}
The cumulative eigenvalue (CE) plot compares the variation components, i.e. the eigenvalues of the correlation matrix. These eigenvalues represent how much information is withheld in each component in terms of explained variance. Thus, by comparing cumulative sums of variance components, it is possible to obtain a detailed picture of how the two datasets differ in terms of what components are the most informative. In order to investigate whether the same proportion of the total variation can be described by the same number of principal components in the two datasets, we plot a piecewise linear curve connecting the points
\begin{align*}
&(0,0), \;
(\lambda_1,\lambda_{11}-\lambda_{21}), \;
(\lambda_1 + \lambda_2,\lambda_{11}+\lambda_{12}-\lambda_{21}-\lambda_{22}), \;
\ldots, \\
&\bigg( \sum_{j=1}^d \lambda_j, \sum_{j=1}^d \lambda_{1j} - \sum_{j=1}^d \lambda_{2j} \bigg)
\end{align*}
Note that due to the standardization, the last point will always be equal to $(d,0)$. Thus, this curve will begin and end at the x-axis. And the larger excursions it makes away from the x-axis, the less alike the cumulative variation components for the two datasets are. Moreover, a positive cumulative difference implies that dataset 1 holds more information in the first components than dataset 2 does. Two examples of cumulative eigenvalue plots are available in Figure \ref{fig:plot.simCE}.

\begin{figure}[!h]
\caption{[FIGURE 2 HERE]}
\label{plot.simCE}
% CE plots for the simulated datasets. Dataset A (with no differences in data structure) in top panel, and dataset B (where observations stem from different models) in bottom panel. Plots are annotated with the $p$-values of the Kolmogorov-Smirnov (KS) and the Cram\'er-von Mises (CvM) tests of the null hypothesis of no difference in data structures. The bold, black lines illustrate the observed cumulative eigenvalue differences, while the shaded areas are 95 \% confidence bands under the null hypothesis of no difference in data structures.
\end{figure}

@@

\begin{figure}[!h]
\caption{[FIGURE 3 HERE]}
% Angle plots for the simulated datasets. Dataset A in top panel and dataset B in bottom panel. Blue arrows show the principal components of the observations in group 1 decomposed in the coordinate system of the principal components of group 2, while the red arrows illustrate the reverse.
\label{plot.simAngle}
\end{figure}

@@

\begin{figure}[!h]
\caption{[FIGURE 4 HERE]}
% Chroma plots for the simulated datasets. Dataset A in top panel and dataset B in bottom panel. Each bar represents a principal component and the bars are annotated with their cumulative percentages of explained variance.
\label{plot.simChroma}
\end{figure}

@@

The ESS 2012 data contain a total of 626 variables collected from 54673 citizens of 29 countries. Here, we will only work with a subset of 35 questionnaire items that are all related to psychological well-being. These 35 items can be divided into 6 distinct scales, namely \textit{Evaluative wellbeing}, \textit{Emotional wellbeing}, \textit{Functioning}, \textit{Vitality}, \textit{Community wellbeing} and \textit{Supportive relationships}. More details on these scales can be found in \cite{ESStopline5} and the relationship between questionnaire items and scales is summarized in the supplementray material. We represent each of the scales by a single variable, which is calculated as the average score within the items related to that variable and scaled such that it takes a value between 0 and 10. For simplicity, we use only complete cases for this construction and thus exclude all participants that did not answer all the 35 questionnaire items used below. This corresponds to approximately 9\% of the observations in the Danish sample, 20\% in the Bulgarian sample and only 6\% in the Swedish sample. All in all, we have $n_{DK} = 1498$ complete case observations in the Danish sample, $n_{BG} = 1798$ observations in the Bulgarian sample and $n_{SE} = 1736$ Swedish observations. Table~\ref{tableDistr} summarizes the marginal distributions of the six dimensions of psychological well-being, stratified by country.

\begin{table}[!ht]
\centering
\begin{tabular}{lccc}
\hline
  & Denmark & Bulgaria & Sweden \\
\hline
Evaluative wellbeing    & 8.75 (8.00, 9.50) & 5.00 (3.50, 7.00) & 8.00 (7.00, 9.00) \\
Emotional wellbeing     & 8.33 (7.22, 8.89) & 6.67 (5.00, 7.78) & 7.78 (6.67, 8.89) \\
Functioning             & 7.57 (6.93, 8.21) & 6.68 (5.50, 7.68) & 7.04 (6.39, 7.68) \\
Vitality                & 7.50 (6.67, 8.33) & 7.50 (5.83, 8.33) & 7.50 (6.67, 9.17) \\
Community wellbeing     & 6.77 (5.83 ,7.57) & 4.67 (3.70, 5.70) & 6.57 (5.66, 7.37) \\
Supportive relationships& 8.25 (7.42, 8.92) & 7.25 (6.17, 8.08) & 8.25 (7.42, 8.75) \\
\hline
\end{tabular}
\caption{Median values of each of the six dimensions of psychological well-being, stratified by country. 25 and 75 percentiles are listed in parentheses. Note that the scales are constructed such that they all run from 0-10.}
\label{tableDistr}
\end{table}

@@

\begin{figure}[!h]
\caption{[FIGURE 5 HERE]}
% CE plots (top panel) and angle plots (bottom panel) comparing Danish and Bulgarian data on psychological well-being. The CE plot is annotated with the $p$-values of the Kolmogorov-Smirnov and the Cram\'er-von Mises tests of the hypothesis of no difference in data structures. In the angle plot, the blue arrows show the principal components of the Bulgarian dataset decomposed in the coordinate system of the principal components of the Danish dataset, while the red arrows illustrate the reverse.
\label{plotBG.cehair}
\end{figure}

@@

\begin{figure}[!h]
\caption{[FIGURE 6 HERE]}
% Chroma plot for comparing Danish and Bulgarian data on psychological well-being. A chroma plot comparing the 2nd, 3rd and 4th principal components of the Bulgiarian- and Danish psychological well-being data. The component-bars are annotated with their relative variance contributions.
\label{plotBG.pancake}
\end{figure}

All in all, we find that psychological well-being does \textit{not} seem to be the same concept in Bulgaria and Denmark. The two countries disagree both in how many dimensions are needed to capture the most important parts of the concept (as illustrated by the differences in eigenvalues) and in how these dimensions are then weighted among the 6 scales (as illustrated by the angle- and chroma plots). In Bulgaria, interpersonal features seem to be more informative of psychological well-being, whereas in Denmark, individual characteristic play a relatively larger role, corresponding well with the previous findings mentioned above. Thus, the datasets are fundamentally different and we should therefore be wary about combining them in a joint analysis, which was also the conclusion of the ESS authors, though based on country-level aggregated statistics \cite{ESStopline5}. Moreover, this also implies that two countries cannot be ranked in terms of which country is "the most happy", at least not by referring to psychological well-being dimensions such as those encountered here.

@@

\begin{figure}[!h]
\caption{[FIGURE 7 HERE]}
% A CE plot (top panel) and an angle plot (bottom panel) comparing Danish and Swedish data on psychological well-being. The blue arrows show the principal components of the Danish dataset decomposed in the coordinate system of the principal components of the Swedish dataset, and the red arrows illustrate the reverse.
\label{plotSE.cehair}
\end{figure}

@@

\begin{figure}[!h]
\caption{[FIGURE 8 HERE]}
% Chroma plot comparing the loading patterns of the Danish and the Swedish subsamples. For each component, the bar is annotated with its cumulative variance contribution, that is, how much variance can be explained by having information of this and the preceding components.
\label{plotSE.pancake}
\end{figure}

>>>>

\documentclass[a4paper,12pt]{article}

\usepackage{geometry}
\geometry{margin=1in}

\usepackage[english]{babel}
\usepackage{authblk}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathrsfs}
%\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
\usepackage{apalike}
\usepackage[utf8x]{inputenc}
%\usepackage{nameref,hyperref}
% line numbers
%\usepackage[right]{lineno}
% ligatures disabled
\usepackage{microtype}
\usepackage{multirow}
\usepackage[table]{xcolor}
\definecolor{lightgray}{HTML}{e6e6e6}

%\DisableLigatures[f]{encoding = *, family = * }
% color can be used to apply background shading to table cells only
\usepackage[table]{xcolor}
% array package and thick rules for tables
\usepackage{array}
% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

@@

\def\volumeyear{2018}

@@

\author[1]{}
%\author[1]{Anne Helby Petersen}
%\author[1]{Bo Markussen}
%\author[3]{Karl Bang Christensen}
%\affil[1]{Department of Public Health, University of Copenhagen, Copenhagen, Denmark}
%\affil[2]{Department of Mathematical Sciences, University of Copenhagen, Copenhagen, Denmark}
\date{}
\maketitle

@@

@@

\begin{abstract}
Datasets are sometimes divided into distinct subsets, e.g. due to multi-center sampling, or to variations in instruments, questionnaire item ordering or mode of administration, and the data analyst then needs to assess whether a joint analysis is meaningful. The Principal Component Analysis-based Data Structure Comparisons (PCADSC) tools are three new non-parametric, visual diagnostic tools for investigating differences in structure for two subsets of a dataset through covariance matrix comparisons by use of principal component analysis. The PCADCS tools are demonstrated in a data example using European Social Survey data on psychological well-being in three countries, Denmark, Sweden and Bulgaria. The data structures are found to be different in Denmark and Bulgaria, and thus a comparison of e.g. mean psychological well-being scores is not meaningful. However, when comparing Denmark and Sweden, very similar data structures, and thus comparable concepts of well-being, are found. Therefore, inter-country comparisons are warranted for these countries.
\end{abstract}

@@

@@

Data comparability is a classical topic in statistics and a recurring one in psychometrics. Often, psychometric data will be collected in such a way that it is essentially divided into several subsets whose comparability needs to be assessed empirically. This happens e.g. when data are collected across several centers (or countries) or when different versions of an instrument, the mode of administration or the order of survey questionnaire items are applied. All of these practices are essential parts of the psychometric methodology, as it allows us to simultaneously build on existing methods and answer new, empirical questions. However, if we do not address the question of comparability among the data subsets arriving from e.g. different countries, we risk conducting analyses whose most fundamental assumptions are not satisfied.

Sophisticated methods for addressing this question are available when we are willing to assume a statistical model. But this places the effort of assessing data comparability very late in the data analysis process and it makes the comparability assessment ad-hoc, as it essentially relies on the appropriateness of the modeling choices, which again depends on the structure of the data. The use of (parametric) models does thus not constitute a general data structure comparison method, but rather a fitted-model comparison method. It addresses the interplay between the model and the data, not the data alone. However, useful tools for initial, exploratory investigations of data comparability that help researchers identify potential problems or challenges early in the study design development are mostly absent. What is needed are procedures that compares differences in data structures in two subsets of a dataset without assuming neither directional nor hierarchical relationships between the variables. This paper suggests procedures for linear data structures Simple methods like variable-by-variable tests of distributional differences suffer from the drawback that they only address marginal differences and ignore the interplay between variables, but entry-by-entry comparisons of two empirical correlation matrices quickly become unmanageable as the number of variables increase.

@@

We propose three visual tools for this task, which we will refer to collectively as Principal Component Analysis-based Data Structure Comparisons (PCADSC). These methods use principal component decomposition of the empirical covariance matrix in the two subsets to create intuitive visualizations of data structure differences. The proposed tools are implemented in the \texttt{R} package \texttt{PCADSC} \cite{PCADSC}.

@@

@@

PCA qualifies as an appealing first step in structural comparisons of two datasets containing the same variables, and especially the loadings and variance components are meaningful and interesting objects to compare across such different datasets. We consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ and let $\bar{x} = {(\bar{x}_1, ..., \bar{x}_d)}^\top = \frac{1}{n} \sum_{i=1}^n x_n$ and let $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) {(x_i-\bar{x})}^\top \in \RR^{d \times d}$ denote their averages and the empirical covariance matrix of the full data matrix, respectively. A brief introdution to CFA is included in the supplementary material.

Usually, in PCA, the main interest lies in the \textit{scores}, i.e. the coordinates $\eta_j^\top (x_i - \bar{x})$ of the projections observations onto the loadings, but the variation components and the accompanying loadings also describe aspects of the data. If two different datasets with the same variables, but different samples of observations, have similar loadings, then the variables can be argued to measure the same underlying quantities in both datasets. Similar loadings can occur in situations where the two sets of scores are different, e.g.\ if the two datasets come from two different populations of subjects. On the other hand, if the loading patterns are different in the two datasets, then this indicates that the variable interplay differs in the two data situations, and hence it would be criticizable to use these variables for comparisons across the two datasets.

The tools presented in this paper are all based on comparing the PCA results across two different datasets that contain the same variables. We denote these two datasets by $X_1$ and $X_2$, respectively, with $X_1$ consisting of $n_1$ observations of $d$ variables $x_{11},\dotsc,x_{1 n_1} \in \RR^d$, and similarly, $X_2$ consisting of $n_2$ observations of the same $d$ variables, $x_{21},\dotsc,x_{2 n_2} \in \RR^d$. %
Furthermore, let $X$ be the combined dataset consisting of all $n=n_1+n_2$ observations. For these %two
three datasets, we complete the following steps:
\begin{enumerate}
\item Within each dataset, standardize each of the variables to have mean zero and unit standard deviation. Let $\tilde{x}_{ij} \in \RR^d$ for $i=1,2$ and $j=1,\dotsc,n_i$ and $\tilde{x}_j \in \RR^d$ for $j=1,\dotsc,n$ be the standardized datasets.
\item Form the principal component analyses
\begin{align*}
S_i &= \frac{1}{n_i-1} \sum_{j=1}^{n_i} \tilde{x}_{ij} \tilde{x}_{ij}^\top = \sum_{k=1}^d \lambda_{ik} \eta_{ik} \eta_{ik}^\top \quad \text{for $i=1,2$,} \\
S &= \frac{1}{n-1} \sum_{j=1}^n \tilde{x}_j \tilde{x}_j^\top = \sum_{k=1}^d \lambda_k \eta_k \eta_k^\top,
\end{align*}
thereby obtaining loadings $\eta_{1k}, \eta_{2k}, \eta_k$ and variance components $\lambda_{1k}, \lambda_{2k}, \lambda_k$ for $k=1,\dotsc,d$.
\end{enumerate}
The hereby obtained PCA decompositions of the correlation matrices can then be compared. We present three diagnostic plots that are designed to shine a light on different types and levels of data structure differences. These three plots are:
\begin{description}
\item[The CE plot:] The CE (cumulative eigenvalue) plot can be used to illustrate differences in variance components, that is, in the relative importance of the directions identified by the PCA. The CE plot is accompanied by two permutation-tests, testing the hypothesis of no difference in variance components.
\item[The angle plot:] The angle plot compares both loadings and variance components at once and it can be used to understand the information loss if the data structure of one dataset is superimposed on the other, thereby revealing which principal components (i.e. loading and variance component pairs) are most similar and most different across the two datasets
\item[The chroma plot:] The chroma plot is primarily an illustration of the loading patterns and it targets the question of how the roles of the original variables are different between the two datasets, thus leading the data structure comparison question back to its original, empirical context.
\end{description}
For the deepest understanding of the data structure differences in two datasets, we suggest using all three steps in the above order.

Note that the standardization implies that the diagonal elements of $S$, $S_1$, and $S_2$ all equal 1, and thus also that
\begin{equation*}
\sum_{k=1}^d \lambda_k = \sum_{k=1}^d \lambda_{1k} = \sum_{k=1}^d \lambda_{2k} =  d.
\end{equation*}
This identity will simplify some expressions below.

@@

In \textit{Model 1} the observed variables $v_3$ and $v_4$ are caused by a common, unobserved variable $X_1$. Further details on the models and the simulated data are available in the supplementary material, including covariance matrices for the two models (S1). Further details about data simulation is also available in the supplementary material. In dataset A, all observations stem from the same, underlying data structure and thus, the PCADSC plots should point towards no notable differences in data structures. In dataset B, on the other hand, there are two fundamentally different, underlying data structures and therefore, the PCADSC plots should illustrate this lack of homogeneity in data structures.

\begin{figure}[h]
\caption{[FIGURE 1 ABOUT HERE]. Graphical representations of models used for simulating example data. Square nodes represent observed variables, while rounded nodes represent latent variables. Arrows are used to illustrate causal links.}
\label{fig:simGraphs}
\end{figure}

\subsubsection*{The cumulative eigenvalue plot}
The cumulative eigenvalue (CE) plot compares the variation components, i.e. the eigenvalues of the correlation matrix. These eigenvalues represent how much information is withheld in each component in terms of explained variance. Thus, by comparing cumulative sums of variance components, it is possible to obtain a detailed picture of how the two datasets differ in terms of what components are the most informative. In order to investigate whether the same proportion of the total variation can be described by the same number of principal components in the two datasets, we plot a piecewise linear curve connecting the points
\begin{align*}
&(0,0), \;
(\lambda_1,\lambda_{11}-\lambda_{21}), \;
(\lambda_1 + \lambda_2,\lambda_{11}+\lambda_{12}-\lambda_{21}-\lambda_{22}), \;
\ldots, \\
&\bigg( \sum_{j=1}^d \lambda_j, \sum_{j=1}^d \lambda_{1j} - \sum_{j=1}^d \lambda_{2j} \bigg)
\end{align*}
Note that due to the standardization, the last point will always be equal to $(d,0)$. Thus, this curve will begin and end at the x-axis. And the larger excursions it makes away from the x-axis, the less alike the cumulative variation components for the two datasets are. Moreover, a positive cumulative difference implies that dataset 1 holds more information in the first components than dataset 2 does. Two examples of cumulative eigenvalue plots are available in Figure \ref{plot.simCE}.

\begin{figure}[h]
\caption{[FIGURE 2 ABOUT HERE] CE plots for the simulated datasets. Dataset A (with no differences in data structure) in top panel, and dataset B (where observations stem from different models) in bottom panel. Plots are annotated with the $p$-values of the Kolmogorov-Smirnov (KS) and the Cram\'er-von Mises (CvM) tests of the null hypothesis of no difference in data structures. The bold, black lines illustrate the observed cumulative eigenvalue differences, while the shaded areas are 95 \% confidence bands under the null hypothesis of no difference in data structures.}
\label{plot.simCE} 
\end{figure}

@@

\begin{figure}[h]
\caption{[FIGURE 3 ABOUT HERE] Angle plots for the simulated datasets. Dataset A in top panel and dataset B in bottom panel. Blue arrows show the principal components of the observations in group 1 decomposed in the coordinate system of the principal components of group 2, while the red arrows illustrate the reverse.}
\label{plot.simAngle}
\end{figure}

@@

\begin{figure}[h]
\caption{[FIGURE 4 ABOUT HERE] Chroma plots for the simulated datasets. Dataset A in top panel and dataset B in bottom panel. Each bar represents a principal component and the bars are annotated with their cumulative percentages of explained variance.}
\label{plot.simChroma}
\end{figure}

@@

The ESS 2012 data contain a total of 626 variables collected from 54673 citizens of 29 countries. Here, we will only work with a subset of 35 questionnaire items that are all related to psychological well-being. These 35 items can be divided into 6 distinct scales, namely \textit{Evaluative wellbeing}, \textit{Emotional wellbeing}, \textit{Functioning}, \textit{Vitality}, \textit{Community wellbeing} and \textit{Supportive relationships}. More details on these scales can be found in \cite{ESStopline5} and the relationship between questionnaire items and scales is summarized in the supplementray material (S2). We represent each of the scales by a single variable, which is calculated as the average score within the items related to that variable and scaled such that it takes a value between 0 and 10. For simplicity, we use only complete cases for this construction and thus exclude all participants that did not answer all the 35 questionnaire items used below. This corresponds to approximately 9\% of the observations in the Danish sample, 20\% in the Bulgarian sample and only 6\% in the Swedish sample. All in all, we have $n_{DK} = 1498$ complete case observations in the Danish sample, $n_{BG} = 1798$ observations in the Bulgarian sample and $n_{SE} = 1736$ Swedish observations. Table~\ref{tableDistr} summarizes the marginal distributions of the six dimensions of psychological well-being, stratified by country.

\begin{table}[ht]
\centering
\caption{\textbf{Median values of each of the six dimensions of psychological well-being, stratified by country.} 25 and 75 percentiles are listed in parentheses. The scales are constructed such that they all run from 0-10.} 
\label{tableDistr}

\begin{tabular}{lccc}
\hline
  & Denmark & Bulgaria & Sweden \\
\hline
Evaluative wellbeing    & 8.75 (8.00, 9.50) & 5.00 (3.50, 7.00) & 8.00 (7.00, 9.00) \\
Emotional wellbeing     & 8.33 (7.22, 8.89) & 6.67 (5.00, 7.78) & 7.78 (6.67, 8.89) \\
Functioning             & 7.57 (6.93, 8.21) & 6.68 (5.50, 7.68) & 7.04 (6.39, 7.68) \\
Vitality                & 7.50 (6.67, 8.33) & 7.50 (5.83, 8.33) & 7.50 (6.67, 9.17) \\
Community wellbeing     & 6.77 (5.83 ,7.57) & 4.67 (3.70, 5.70) & 6.57 (5.66, 7.37) \\
Supportive relationships& 8.25 (7.42, 8.92) & 7.25 (6.17, 8.08) & 8.25 (7.42, 8.75) \\
\hline
\end{tabular} 
\end{table}

@@

\begin{figure}[h]
\caption{[FIGURE 5 ABOUT HERE] CE plots (top panel) and angle plots (bottom panel) comparing Danish and Bulgarian data on psychological well-being. The CE plot is annotated with the $p$-values of the Kolmogorov-Smirnov and the Cram\'er-von Mises tests of the hypothesis of no difference in data structures. In the angle plot, the blue arrows show the principal components of the Bulgarian dataset decomposed in the coordinate system of the principal components of the Danish dataset, while the red arrows illustrate the reverse.}
\label{plotBG.cehair}
\end{figure}

@@

\begin{figure}[h]
\caption{[FIGURE 6 ABOUT HERE] Chroma plot for comparing Danish and Bulgarian data on psychological well-being. A chroma plot comparing the 2nd, 3rd and 4th principal components of the Bulgarian- and Danish psychological well-being data. The component-bars are annotated with their relative variance contributions. }
\label{plotBG.pancake}
\end{figure}

All in all, we find that psychological well-being does \textit{not} seem to be the same concept in Bulgaria and Denmark. The two countries disagree both in how many dimensions are needed to capture the most important parts of the concept (as illustrated by the differences in eigenvalues) and in how these dimensions are then weighted among the 6 scales (as illustrated by the angle- and chroma plots). In Bulgaria, interpersonal features seem to be more informative of psychological well-being, whereas in Denmark, individual characteristics play a relatively larger role, corresponding well with the previous findings mentioned above. Thus, the datasets are fundamentally different and we should therefore be wary about combining them in a joint analysis, which was also the conclusion of the ESS authors, though based on country-level aggregated statistics \cite{ESStopline5}. Moreover, this also implies that two countries cannot be ranked in terms of which country is "the most happy", at least not by referring to psychological well-being dimensions such as those encountered here.

@@

\begin{figure}[h]
\caption{[FIGURE 7 ABOUT HERE] A CE plot (top panel) and an angle plot (bottom panel) comparing Danish and Swedish data on psychological well-being. The blue arrows show the principal components of the Danish dataset decomposed in the coordinate system of the principal components of the Swedish dataset, and the red arrows illustrate the reverse.}
\label{plotSE.cehair}
\end{figure}

@@

\begin{figure}[h]
\caption{[FIGURE 8 ABOUT HERE] Chroma plot comparing the loading patterns of the Danish and the Swedish subsamples. For each component, the bar is annotated with its cumulative variance contribution, that is, how much variance can be explained by having information of this and the preceding components.}
\label{plotSE.pancake}
\end{figure}
